---
title: "dacss604_final_project"
execute: 
  echo: false
---

# Polarization, Division: The New American Norm?

\[Topic Sentence Here\]. To even casual observers, polarization and division seem to characterize American politics. Bitter political cleavages have been formed along the lines of several factors: political ideologies (**Source**), age groups (**Source**), racial/ethnic identities (**Source**), and more. One divide stands out as particularly stark: geography. Rural and Urban politics are particularly polarized. Americans living in different types of geographic settings have opposing ideas on political issues (**Source**). *Give examples polarization*.

*Paragraph about geographic polarization/division in general. Cite studies of issue in Italy, France.*

Political division has wide-reaching negative effects on both individuals and groups (**Sources**). *Give examples of such negative effects*. Predictions have been made that such polarization may deepen in coming years. This division and polarization is established and worrisome, but much of the research examines politics at the national level: what about politics at the state level? This project contributes to the investigation of this question.

# This Project

This project examines the role of geographic division in Massachusetts state politics. Far less research has been dedicated to this topic than politics at the national level. However, state governments have significant power in the US. Therefore, it bears worth to examine how the important phenomenon of political division may be affecting governance at the state level. Each state has its own politics and issues, therefore it would be a monumental task to examine all states in one project. To approach the matter with appropriate depth, this project examines politics in one state only, The Bay State: The Commonwealth of Massachusetts.

Massachusetts' state government has a similar structure to the US federal government: An executive branch, responsible for the enforcement of laws, headed by a single executive (the governor) with numerous agencies; A bicameral legislative branch (having two distinct chambers) charged with the creation of new laws; and a judicial branch, composed of a system of courts that interpret the laws. This project investigates the question of geographic division in Massachusetts politics by examining the legislative branch specifically.

This project uses a mixed methods (utilizing both quantitative and qualitative methods) approach to study political cooperation in the Massachusetts legislature. The quantitative portion examines the Massachusetts House of Representatives, Massachusetts' equivalent to the US' House of Representatives. The House of Representatives consists of elected legislators that represent the different geographic districts of Massachusetts. The qualitative portion studies the other chamber of Massachusetts' legislature, the Massachusetts Senate.

## Quantitative: Network Analysis of Massachusetts House of Representatives, 2021–2022

The quantitative portion of the project examines political cooperation in Massachusetts' House of Representatives during the 192nd General Court (the name of Massachusetts' legislature). This legislative body was in session throughout 2021 and 2022. The primary method of investigation was network analysis of this body. *Brief Primer on network theory/analysis.*

In this project, nodes represent individual representatives (legislators) in the 192nd General Court, and edges represent the cosponsoring of a bill. For example, if Representative A and Representative B both sponsor bill X, then there would be an edge between those two representatives. In this project, edges are **weighted**. If edges are weighted, it means that they have a value associated with them, representing some measurement. In this project, edge weight represents how many bills two nodes cosponsored. For instance, in the previous scenario, Representative A and Representative B cosponsored one bill, therefore the weight of the edge connecting them would be one. In another example, Representative B and Representative C sponsored three of the same bills, therefore the weight of the edge between them would be three. The data consisted of 87 nodes (individual representatives) and 50 bills that formed edges (connections) between those representatives. All bills were introduced during the 192nd General Court.

### The Center of the Network

This analysis yielded multiple insights into the nature of political cooperation in the Massachusetts legislature. The graph below (Figure 1) shows the 10 nodes with the highest **degree**. In this case, a node's degree is equal to the sum of the weights of all it's different edges. For example, if Node A has 3 edges, all with weights of 2, then Node A has a degree of 6. Examining the nodes with the highest degrees is useful because it shows us which nodes are the most connected to other nodes in the network, i.e., which representatives are the most collaborative in this sample. Edges' weights are denoted by their thickness: the thicker a connection between two nodes, the more bills those two nodes cosponsored together.

With the shape of the nodes representing geographic-setting-type (circle = rural district, square = urban district), we can see that the most collaborative representatives consist of a heterogeneous mix of urban-representing and rural-representing legislators. 6 of these nodes represent urban districts, 4 represent rural districts, and each of these nodes are heavily connected to one another. This paints a picture of rural- and urban-representing legislators collaborating by sponsoring the same bills, denoting a lack of division and polarization.

```{r load_packages, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
#loading packages

library(tidyverse)
library(igraph)
library(statnet)
library(readxl)
library(stringr)
library(stringi)
library(GGally)
library(blockmodeling)
library(ggplot2)
library(knitr)
library(kableExtra)
```

```{r load_data}
#loading data

RepsList <- read_excel("data/RepsList.xlsx")

DistsPops <- read_excel("data/HouseDistsPops.xlsx", sheet = "HouseDistsPops",
                       col_names = c("district", "population"))

BillsList <- read_excel("data/BillsList2.xlsx",
                        sheet = "Sheet4",
                        col_names = T)

RepsNames <- read_excel("data/BillsList2.xlsx",
                        sheet = "Sheet5")

DistsArea <- read_excel("data/dists.xlsx",
           col_names = T)
```

```{r clean_distsPops}
#Cleaning DistsPops df

DistsPops <- DistsPops %>%
  filter(str_detect(district, "District"),
         str_detect(district, "Total")) %>%
  mutate(district = str_remove(district, "District"),
         district = str_remove(district, "Total"))

DistsPops[61, 1] = "MIDDLE03"
DistsPops[51, 1] = "HAMPDEN08"
DistsPops[6, 1] = "BDN01"
DistsPops[51,2] = "43280"
DistsPops[61, 2] = "43677"
DistsPops$district <- trimws(DistsPops$district)

DistsPops <- DistsPops %>%
  mutate(district = stri_reverse(district),
         district = str_replace(district, pattern = "(.{2})(.*)", replacement = "\\1 \\2"),
         district = stri_reverse(district),
         district = gsub(" 0", " ", district), #use regular expression to replace only "0" preceded by " "
         district = gsub("BARN", "barnstable", district),
         district = gsub("BERK", "berkshire", district),
         district = gsub("BRISTOL", "bristol", district),
         district = gsub("ESSEX", "essex", district),
         district = gsub("FRANK", "franklin", district),
         district = gsub("HAMPDEN", "hampden", district),
         district = gsub("HAMPSHIRE", "hampshire", district),
         district = gsub("MIDDLE", "middlesex", district),
         district = gsub("NORFOLK", "norfolk", district),
         district = gsub("PLY", "plymouth", district),
         district = gsub("SUFFOLK", "suffolk", district),
         district = gsub("WOR", "worcester", district))

DistsPops[6, 1] = "barnstable, dukes and nantucket"
```

```{r}
#cleaning DistsArea df

DistsArea <- DistsArea %>%
  rename(rep = REP,
         repDist = 'Rep Dist',
         repDistNum = Repdistnum,
         area = Area) %>%
  subset(select = c(repDist,
                    area)
         ) %>%
  mutate(repDist = gsub("(.*) (.*)", "\\2 \\1", repDist),
         repDist = str_to_lower(repDist),
         repDist = gsub('.{2}$', '', repDist)) %>%
  drop_na(repDist)

DistsArea[155, 1] = "barnstable, dukes and nantucket"
```

```{r clean_repsList}
#cleaning RepsList df

RepsList <- RepsList %>%
  mutate(District = gsub("(.*) (.*)", "\\2 \\1", District),
         District = str_to_lower(District),
         District = substr(District,
                           1,
                           nchar(RepsList$District)-2),
         name = paste(`First Name`, `Last Name`, sep = " ")
         ) %>%
  relocate(name, .after = `Last Name`) %>%
  subset(select = c(name, District, Party))

RepsList <- left_join(RepsList, RepsNames, by = join_by(name == name)) %>%
  relocate(id, .before = name) %>%
  # subset(select = c(id, District, Party)) %>%
  drop_na(id)

RepsList[28, 3] = "barnstable, dukes and nantucket"

RepsList <- RepsList %>%
  left_join(DistsArea,
            by = join_by(District == repDist)) %>%
  left_join(DistsPops,
            by = join_by(District == district)) %>%
  mutate(popDens = as.integer(population)/as.integer(area),
         classification = case_when(
           popDens >= 1000 ~ "urban",
           popDens < 1000 ~ "rural"
         ))
```

```{r wrangle_billsListEdgeList}
#cleaning and transforming BillsList df

#removing bills that have only 1 sponsor
BillsList <- BillsList %>%
  drop_na('Sponsor 2')

#creating rough edgelist
result <- stack(table(unlist(apply(BillsList[-1], 1, function(x) {
                 vec <- na.omit(x)
                 if(length(vec) < 2) return(NULL)
                  combn(vec, 2, function(y) paste0(sort(y), collapse = '—'))
            }))))[2:1]

BillsList.el <- as.matrix(separate(result, ind, c('nodeA', 'nodeB'), sep = '—'))

BillsList.el.df <- as.data.frame(BillsList.el)

BillsList.el.df <- BillsList.el.df %>%
  left_join(RepsList,
            by = join_by(nodeA == name)) %>%
  rename(nodeA.id = id) %>%
  relocate(nodeA.id,
           .before = values) %>%
  subset(select = c(nodeA,
                    nodeB,
                    nodeA.id,
                    values)
         ) %>%
  left_join(RepsList,
            by = join_by(nodeB == name)) %>%
  rename(nodeB.id = id,
         weight = values) %>%
  relocate(nodeB.id,
           .before = weight) %>%
  subset(select = -c(nodeA,
                     nodeB,
                     District,
                     Party,
                     area,
                     population,
                     popDens,
                     classification)) %>%
  drop_na(nodeA.id,
          nodeB.id) %>%
  mutate(weight = as.integer(weight))

BillsList.el <- as.matrix(BillsList.el.df)
```

```{r create_adjMatrix}
#creating adjacency matrix from edgelist

g1 <- graph_from_data_frame(BillsList.el.df)
g1.1 <- as_adjacency_matrix(g1,sparse = FALSE, attr = "weight")

#symmetrizing the matrix
g1.2 <- (g1.1) + t(g1.1)

#Vertices too small
#curved edges too small
```

```{r create_adjMatrix_p2}
#switching to iGraph method
#rough visualization of the network
pol_net <- graph_from_data_frame(d = BillsList.el,
                                 directed = F)


#making nodes smaller
#removing node labels
#rough visualization of the network

#Conclusion: Whole network has too many nodes & Edges for meaningful visualizations
#Next Steps: select top 5, 10, 15 nodes in terms of edges and visualize only them
```

```{r create_repsList}
#ranking nodes based on degree
#first, find degree of nodes
deg <- igraph::degree(pol_net)

#next create dataframe with col1 <- node id, col 2 <- node's degree
deg1 <- deg %>%
  map(~as_tibble(.)) %>%
  bind_rows(.id = "index") %>%
  mutate(index = as.integer(index)) %>%
  rename("degree" = "value")

#next joining previously made df to node attribute list to list each node's degree as an attribute
RepsList <- left_join(RepsList,
          deg1,
          by = join_by("id" == "index"))

#dropping na's in rep's degrees
RepsList <- RepsList %>%
  drop_na(degree)

#next ranking reps by their degree; Variations w/ top 5, 10, 15 and 20 nodes
RepsList_deg5 <- head(RepsList %>%
  arrange(desc(degree)),
  n = 5)

RepsList_deg10 <- head(RepsList %>%
  arrange(desc(degree)),
  n = 10)

RepsList_deg15 <- head(RepsList %>%
  arrange(desc(degree)),
  n = 15)

RepsList_deg20 <- head(RepsList %>%
  arrange(desc(degree)),
  n = 20)
```

```{r create_edgeList}
#creating edgelists containing each of top 5, 10, 15, 20 nodes in terms of degree
BillsList.el.df_deg5 <- BillsList.el.df %>%
  filter(nodeA.id %in% c(4, 36, 55, 40, 26) &
           nodeB.id %in% c(4, 36, 55, 40, 26))

BillsList.el.df_deg10 <- BillsList.el.df %>%
  filter(nodeA.id %in% c(4, 36, 55, 40, 26, 63, 35, 62, 2, 71) &
           nodeB.id %in% c(4, 36, 55, 40, 26, 63, 35, 62, 2, 71))

BillsList.el.df_deg15 <- BillsList.el.df %>%
  filter(nodeA.id %in% c(4, 36, 55, 40, 26, 63, 35, 62, 2, 71, 30, 77, 31, 74, 18) &
           nodeB.id %in% c(4, 36, 55, 40, 26, 63, 35, 62, 2, 71, 30, 77, 31, 74, 18))

BillsList.el.df_deg20 <- BillsList.el.df %>%
  filter(nodeA.id %in% c(4, 36, 55, 40, 26, 63, 35, 62, 2, 71, 30, 77, 31, 74, 18, 57, 75, 53, 44, 60) &
           nodeB.id %in% c(4, 36, 55, 40, 26, 63, 35, 62, 2, 71, 30, 77, 31, 74, 18, 57, 75, 53, 44, 60))
```

```{r make_edgeList_networkObject}
#turning each edgelist into a network object

graph_deg5 <- graph_from_data_frame(BillsList.el.df_deg5,
                           directed = FALSE)

graph_deg10 <- graph_from_data_frame(BillsList.el.df_deg10,
                           directed = FALSE)

graph_deg15 <- graph_from_data_frame(BillsList.el.df_deg15,
                           directed = FALSE)

graph_deg20 <- graph_from_data_frame(BillsList.el.df_deg20,
                           directed = FALSE)

#Rough, basic visualizations have been created for each of the new network objects
#next, improving them visually
```

#### Figure 1:

```{r basic_network_viz, results='hide'}
#assigning network visualization features to node attributes

#node color = geographic setting type
#node shape = party

  #top 10 degree nodes
shapes_deg10 <- case_when(RepsList_deg10$classification == "urban" ~ "square",
                  RepsList_deg10$classification == "rural" ~ "circle")
cols_deg10 <- case_when(RepsList_deg10$Party == "Democrat" ~ "blue",
                         RepsList_deg10$Party == "Republican" ~ "red")

par(bg = "grey94") 

  #top 10 degree nodes
V(graph_deg10)$color <- cols_deg10
V(graph_deg10)$shape <- shapes_deg10
net10_plot <- plot(graph_deg10,
                   vertex.size = 0.5*RepsList_deg10$degree,
                   vertex.label.color = "white",
                   vertex.lablel.bg = "black",
                   edge.width = BillsList.el.df_deg10$weight*1,
                   main = "The 10 Most Central (Connected) Nodes",
                   sub = "Heterogenous Composition of Rural and Urban Representatives",
                   curved = 0.2)

net10_plot

legend("topright", 
       legend = c("Democrat", "Republican"), 
       fill = c("blue", "red"), 
       title = "Party",
       border = "black")

legend("bottomleft", 
       legend = c("Urban", "Rural"), 
       pch = c(15, 16),  # Square for Urban, Circle for Rural
       col = c("black", "black"), 
       title = "Geographic Setting",
       pt.bg = c("blue", "red"))
```

```{r create_statnet_objects}
#creating statnet object from symmetrized adjacency matrix
pol_net.stat <- as.network(x = g1.2,
                           directed = F,
                           ignore.eval = F,
                           names.eval = "weight",
                           loops = F,
                           matrix.type = "adjacency")
```

```{r set_node_attributes}
#setting nodal attributes to new statnet network

set.vertex.attribute(pol_net.stat,
                     "full_name",
                     RepsList$name)
set.vertex.attribute(pol_net.stat,
                     "party",
                     RepsList$Party)
set.vertex.attribute(pol_net.stat,
                     "geo_setting",
                     RepsList$classification)
```

```{r struct_equiv_blk_mdl}
get.eigen<-function(net, attr=NULL){
    attr="weight"
    eigen<-sna::evcent(net)
    mat<-as.matrix.network(net, attr=attr)
    diag(mat)<-0
    mat2<-mat%*%mat
    rc<-diag(mat2)/rowSums(mat2)
    dc<-1-rc
    data.frame(name=net%v%"vertex.names",
        eigen=eigen,
        eigen.rc=eigen*rc,
        eigen.dc=eigen*dc)
}
```

```{r create_df_centr_measures}
pol_net.nodes <- data.frame(name=pol_net.stat%v%"vertex.names",
        degree=sna::degree(pol_net.stat,gmode="graph"),
        degree.wt=igraph::strength(pol_net),
        bonpow = igraph::power_centrality(pol_net),
        betweenness=sna::betweenness(pol_net.stat, gmode="graph"),
        close=sna::closeness(pol_net.stat, gmode="graph"),
        constraint=igraph::constraint(pol_net))

pol_net.nodes <- full_join(pol_net.nodes, get.eigen(pol_net.stat), by="name")
```

```{r calc_struc_equiv}
# pol_net.se <- equiv.clust(pol_net.stat,
#                           equiv.fun = "sedist",
#                           method = "hamming", 
#                           mode = "graph")

pol_net.sedist <- blockmodeling::sedist(g1.2)
pol_net.se <- sna::equiv.clust(pol_net.stat, 
                        equiv.dist = pol_net.sedist)
```

```{r calcing_blockmodeling}
# blockmodel and select partitions: 5 partitions
blk_mod1 <- blockmodel(pol_net.stat, pol_net.se, k=5)
# assign block membership to vertex attribute
pol_net.stat%v%"role" <- blk_mod1$block.membership[match(pol_net.stat%v%"vertex.names",
                                                         blk_mod1$plabels)]
```

```{r plotting_network_roles}
#plotting network
block_plot1 <- ggnet2(pol_net.stat,
       node.color = "role",
       node.size = degree(pol_net.stat, gmode = "graph"),
       node.label = F,
       node.shape = "geo_setting",
       node.alpha = 0.7)
```

```{r dendrograms_p1}
#cut height 15

blk_mod2 <- blockmodel(pol_net.stat, pol_net.se, h=15)
pol_net.stat%v%"role" <- blk_mod2$block.membership[match(pol_net.stat%v%"vertex.names",
                                                         blk_mod2$plabels)]

#visualizing
block_plot2 <- ggnet2(pol_net.stat,
       node.color = "role",
       node.size = degree(pol_net.stat, gmode = "graph"),
       node.label = F,
       node.shape = "geo_setting",
       node.alpha = 0.7)
```

```{r dendrograms_p2}
#cut height 10

blk_mod3 <- blockmodel(pol_net.stat, pol_net.se, h = 10)
pol_net.stat%v%"role" <- blk_mod3$block.membership[match(pol_net.stat%v%"vertex.names",
                                                         blk_mod3$plabels)]

#visualizing
block_plot3 <- ggnet2(pol_net.stat,
       node.color = "role",
       node.size = degree(pol_net.stat, gmode = "graph"),
       node.label = F,
       node.shape = "geo_setting",
       node.alpha = 0.7)
```

```{r dendrograms_p3}
#cut height 5

blk_mod4 <- blockmodel(pol_net.stat, pol_net.se, k=10)
pol_net.stat%v%"role" <- blk_mod4$block.membership[match(pol_net.stat%v%"vertex.names",
                                                         blk_mod4$plabels)]

#visualizing
block_plot4 <- ggnet2(pol_net.stat,
       node.color = "role",
       node.size = degree(pol_net.stat, gmode = "graph"),
       node.label = F,
       node.shape = "geo_setting",
       node.alpha = 0.7)
```

```{r}
pol_net.nodes$role <- pol_net.stat%v%"role"
```

```{r dendrograms_p4}
#cut 10 groups

blk_mod5 <- blockmodel(pol_net.stat,
                       pol_net.se,
                       k=5)
pol_net.stat%v%"role" <- blk_mod5$block.membership[match(pol_net.stat%v%"vertex.names",
                                                         blk_mod5$plabels)]

#visualizing
block_plot5 <- ggnet2(pol_net.stat,
       node.color = "role",
       node.size = degree(pol_net.stat, gmode = "graph"),
       node.label = F,
       node.shape = "geo_setting",
       node.alpha = 0.7)
```

```{r node_measures, results='hide'}
nodes_measures <- pol_net.nodes %>%
  select(-name) %>%
  group_by(role) %>%
  mutate(n=n()) %>%
  summarise_all(mean, na.rm = TRUE) %>%
  as.matrix()
```

### Blockmodeling the Network

We can learn more from this network. Another aspect of network theory, **Block Modeling**, can provide more insights about the nature of political collaboration in the Massachusetts legislature. Blockmodeling is a method used to simplify and study complex networks. It works by grouping nodes that have similar patterns of edges together, forming "blocks." Therefore, nodes that are grouped together have similar roles within the networks. By analyzing which blocks different representatives are grouped into, we can learn about patterns of collaboration within this network.

The table below (Figure 2) shows a table that summarizes important information yielded by the blockmodeling analysis done for this project. In this case, the network was divided into 10 blocks. Each row in the table represents a distinct block. The different columns show how many rural-district-representing, urban-district-representing, Democrat, and Republican legislators are in each block. It also shows the ratios of rural:urban representatives and Democrat:Republican representatives in each group.

#### Figure 2:

```{r create_blkmdl_tables}

#creating tables to show block membership

pol_net.nodes <- pol_net.nodes %>%
  mutate(name = as.numeric(name)) %>%
  select(name,
         role)

blk_mdl_tbl <- RepsList %>%
  left_join(pol_net.nodes,
            by = join_by(id == name)) %>%
  group_by(role) %>%
  rename("Group" = "role") %>%
  summarise(Rural = sum(classification == "rural"),
         Urban = sum(classification == "urban"),
         Democrat = sum(Party == "Democrat"),
         Republican = sum(Party == "Republican")) %>%
  mutate("Rural:Urban_Ratio" = Rural/Urban,
         "Democrat:Republican_Ratio" = Democrat/Republican)

blk_mdl_tbl <- kable(blk_mdl_tbl, caption = "Blockmodel Group Statistics", digits = 2, format = "markdown")

blk_mdl_tbl
```

We can see that some groups show a diverse collection of rural and urban representatives (e.g. groups 1, 3, and 4), while others skew more heavily rural or urban (e.g. groups 2, 6, and 7). This shows that in some sections of the legislature, there is high collaboration between legislators representing different geographic types. In other factions, collaboration is more common between legislators that represent the same type of district. This adds nuance to our understanding of the network. We can see that there is certainly a significant level of collaboration between rural and urban representatives—confirming what Figure 1 shows us—but there are also portions of the legislature where such collaboration is less common.

What does this tell us? It tells us that there was a moderate amount of rural-urban collaboration throughout Massachusetts 192nd House of Representatives, however it was concentrated in certain groups of legislators, while other groups displayed less inter-group collaboration. To make an analogy, picture a high-school cafeteria: some portion of the students are upperclassmen, the rest are lowerclassmen. A significant number of the tables in the cafeteria have both upperclassmen and lowerclassmen sitting at them, showing considerable mingling between the two groups. At the same time, there is also a significant number of tables that have mostly (or all) upperclassmen or lowerclassmen. In this cafeteria, inter-group mingling is common, but groups keeping to themselves is also common.

## Qualitative: Interview and Observation of Massachusetts Senate, 2023–2024

The qualitative portion of this project examines political cooperation in Massachusetts' Senate during the 193rd General Court, which was in session throughout 2023 and 2024. The primary methods included in this portion of the project were an interview with a political insider—a senior member of a Massachusetts State Senator's staff—and ethnographic observation of legislative sessions. This portion of the project focused on adding depth of understanding to the patterns observed in the quantitative portion. The interview and observation sessions gave deeper insight into what political cooperation in Massachusetts state politics looks like, and even why we see the patterns that we do.

Unlike the quantitative portion of the project, this portion did not focus on finding broad patterns, rather it was designed to evoke depth and nuance that complement the patterns found in the quantitative portion. These qualitative methods provided rich details about attitudes and experiences of legislators and political insiders in the Massachusetts General Court. The semi-structured interview with a political insider yielded candid information that would be nearly impossible to find within quantitative data, and the ethnographic observation revealed what the nature of political collaboration in Massachusetts' legislature looks like in action. Both methods are detailed below, with a synthesis of their results.

### The Interview

In August 2024, I interviewed a senior staff member ("The Staffer") of a Massachusetts State Senator ("The Senator"). The Staffer is responsible for assisting The Senator with much of their legislative work. The Staffer has candid access to most of The Senator's dealings, and therefore has privileged access to the political operations within the Massachusetts State Senate. Both The Staffer's and The Senator's identities are kept secret for the purposes of confidentiality.

The interview was semi-structured, meaning that there was a loose written script with specific questions and general ideas to explore, but I was free to go off-script and improvise questions and dialogue as I saw fit, which I frequently did. I asked The Staffer what they thought about political collaboration in the Massachusetts Senate: do senators more frequently collaborate, or is division more common? Are their certain issues that are more likely evoke collaboration or division? How much does any present division slow down the process of legislation?

The Staffer described the Senate as being a generally collaborative place, including between rural-representing and urban-representing senators, though with some caveats. The Staffer explained that neither rural vs urban nor democrat vs republican were the primary causes of division in the legislature. In fact, in the Staffer's experience, rural and urban senators much more frequently worked together than they did against one another. In The Staffer's perspective, the primary source of division in the Massachusetts legislature was between the two chambers: The House of Representatives and the Senate. According to the staffer, proceedings within each house were relatively harmonious in most ways, but the two chambers frequently disagreed and failed to collaborate with one another. The Staffer explained that this division was a significant impediment to progress, much more so than any division based on geography.

These results both confirm and deepen the results of the quantitative portion. Once again, collaboration is shown to be common between rural and urban legislators in MA, yet division still exists in some pockets of the General Court. Furthermore, this interview provides a real-time example of the harms of political division. The division between Massachusetts' House of Representatives and Senate slows the pace of legislation, meaning important issues will go unaddressed for greater amounts of time, possibly with significant consequences for the people of MA.

### Observation of Legislative Sessions

The second component of the qualitative portion of this project was consisted of ethnographic observation of several formal sessions of legislation in the Massachusetts House of Representatives during the 193rd General Court (2023–2024). I observed representatives as they formally deliberated on bills by giving speeches that declared their reasons for supporting or opposing the relevant bill. During these speeches, representatives frequently referenced other representatives that had deliberated on the bill, referencing both their arguments, and more personal details. These sessions of observation provided rich detail into how elements of this political network actually interact, and the subjective nature of these interactions. These observations added color and detail to the broad patterns observed in the quantitative portion of the project.

The analysis of these observation sessions consisted of analyzing specific language, sentiments, and themes common to the representatives' speeches. How did representatives speak to one another? Did they usually use more positive and supportive language, or negative and oppositional? Was there a difference in how representatives spoke to/about representatives from different setting types (rural vs urban) vs representatives from the same setting type? These were the kinds of questions which this component of the project was designed to answer.

Overall, when representatives took the initiative to give formal speeches in support or opposition of a bill, the sentiments and languages were highly positive and collaborative, including when representatives were addressing/referencing a representative of the opposite setting type. Sentiments of cooperation were far more common than sentiments of opposition, and rural representatives frequently used language to support urban representatives, and vice versa. The bitter hostility that seems to embody contemporary Federal Politics in the US was noticeably absent.
